Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{ Too \snify specific... }

### 2025-01-10 doc/cicm25-sysdesc/sysdesc.tex
more writing

@@

The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
the domain model for semantic annotations,
as well as re-usable lecture materials such as lecture slides and problems -- mostly from the area of computer science.
The sheer size of the commons is one of the key challenges for authors:
\SMGloM contains definitions of $X$\ednote{TODO} STEM concepts in multiple languages,
created and maintained by over 30 contributors.
Additionally, there are roughly 4900\ednote{
    Includes 440 from UPS, which seems unannotated... Other issues?
    Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
}
semantically annotated lecture slides and 4000 problems.

\paragraph{Overview}
We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
describe and evaluate one concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
The paper ends with a brief conclusion (\cref{sec:conclusion}).
\ednote{rework this paragraph a bit}

@@

\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal
and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But such features only support existing annotations and are, therefore,
of limited use when creating new annotations, which arguably is the
most time consuming aspects of semantic authoring.

How the annotation process can be supported depends on the particular semantic framework,
the specific annotations we want to support, and the semantics embedded in the content commons.
In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers to their definienda.
A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline|A\times B|) into abstract syntax trees based on
semantic macros defined in \sTeX, which can then replace the original formula
(e.g.\ \lstinline|\cart{A,B}| for the cartesian product of sets).
In the case of ambiguity, authors can pick the correct reading in a
graphical user interface that visualizes the different abstract syntax trees.

@@

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
lecture notes; then the session consists of all included files. Other productivity
features include a focus mode that can be used to only annotate a particular word in the
rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
annotation. This reduces the cognitive overhead from switching between different 
words and symbol lists.
% Of course the include management still needs to be done locally.

@@

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that for
effective annotation users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

Finally, the current interface is not well-suited for on-the-fly annotation while
authoring.  For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.

@@

In this paper we have presented the semantic authoring problem using the \sTeX corpus and
the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
from both classical authoring and formal authoring in terms of the necessary system
support. As an example of specialized authoring support, we have presented \snify, a
command-line tool that bridges the gap between formal symbols in \sTeX and

>>>>

% KEY TODOs:
%  * better title
%  * turn overview into combined overview and contribution
%  * Finish tool support section
%  *  -> reduce future work paragraph in conclusion accordingly

@@

The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
the domain model for semantic annotations,
as well as re-usable lecture materials such as lecture slides and problems -- mostly from the area of computer science.
The sheer size of the commons is one of the key challenges for authors:
\SMGloM contains definitions of $X$\ednote{TODO} STEM concepts in multiple languages,
created and maintained by over 30 contributors.
Additionally, there are roughly 4900\ednote{
    Includes 440 from UPS, which seems unannotated...
    Excludes talks as well...
    Other issues?
    Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
}
semantically annotated lecture slides and 4000 problems.

\paragraph{Overview}
We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
describe and evaluate one concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
The paper ends with a brief conclusion (\cref{sec:conclusion}).
\ednote{rework this paragraph a bit -- make it contribution and overview at the same time?}

@@

\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal
and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But such features only support existing annotations and are, therefore,
of limited use when creating new annotations, which arguably is the
most time consuming aspects of semantic authoring.
\ednote{also: SPARQL API}

How the annotation process can be supported depends on the particular semantic framework,
the specific annotations we want to support, and the semantics embedded in the content commons.
In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers to their definienda.
A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on
semantic macros defined in \sTeX, which can then replace the original formula
(e.g.\ \lstinline|\cart{A,B}| for the cartesian product of sets).
In the case of ambiguity, authors can pick the correct reading in a
graphical user interface that visualizes the different abstract syntax trees.

@@

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
lecture notes; then the session consists of all (transitively) included files. Other productivity
features include a focus mode that can be used to only annotate a particular word in the
rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
annotation. This reduces the cognitive overhead from switching between different 
words and symbol lists.
% Of course the include management still needs to be done locally.

@@

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

Finally, the current interface is not well-suited for on-the-fly annotation while
authoring.
For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.
For now, the typical workflow of early adopters is to first write the document (e.g.\ a set of quiz questions) without annotations, and then run \snify to annotate the new content.

@@

In this paper we have presented the semantic authoring problem using the \sTeX corpus and
the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
from both classical authoring and formal authoring in terms of the necessary system
support. As an example of specialized authoring support, we have presented \snify, a
command-line tool that bridges the gap between informal text and formal symbols in \sTeX.
